Estimating Correspondences of Deformable Objects ``In-the-wild"

### 2015-11-05 contents/introduction.tex

\begin{figure}[t!]
    \centering
    \begin{subfigure}[b]{0.09125\textwidth}
            \includegraphics[height=3.25cm]{resources/MotivativeAnnotation/MPII}
            \caption{MPII}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.1125\textwidth}
            \includegraphics[height=3.25cm]{resources/MotivativeAnnotation/FashionPose}
            \caption{Fashion}
    \end{subfigure}
  	\hfill
    \begin{subfigure}[b]{0.125\textwidth}
            \includegraphics[height=3.25cm]{resources/MotivativeAnnotation/FLIC}
            \caption{FLIC}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.125\textwidth}
            \includegraphics[height=3.25cm]{resources/MotivativeAnnotation/BBCPose}
            \caption{BBC Pose}
    \end{subfigure}
    \caption{Visualisation of inconsistent and inaccurate annotations of human pose among different datasets. \emph{Blue} dots denote the original annotations. The arrows and \emph{green} dots show the correct location at which the points should be.
    %Original annotations are shown in blue dots while better annotations shown in green dots.
    }
    \label{fig:wrong_anno}
\end{figure}
\begin{figure}[t!]
    \centering
    \begin{subfigure}[b]{0.065\textwidth}
            \includegraphics[height=2.55cm]{resources/Fig_Intro/intro_2_0}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.065\textwidth}
            \includegraphics[height=2.55cm]{resources/Fig_Intro/intro_2_1}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.065\textwidth}
            \includegraphics[height=2.55cm]{resources/Fig_Intro/intro_2_2}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.085\textwidth}
            \includegraphics[height=2.55cm]{resources/Fig_Intro/intro_1_0}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.085\textwidth}
            \includegraphics[height=2.55cm]{resources/Fig_Intro/intro_1_1}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.085\textwidth}
            \includegraphics[height=2.55cm]{resources/Fig_Intro/intro_1_2}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{0.156\textwidth}
            \includegraphics[height=2.8cm]{resources/Fig_Intro/intro_0_0}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.156\textwidth}
            \includegraphics[height=2.8cm]{resources/Fig_Intro/intro_0_1}
    \end{subfigure}
  	\hfill
    \begin{subfigure}[b]{0.156\textwidth}
            \includegraphics[height=2.8cm]{resources/Fig_Intro/intro_0_2}
    \end{subfigure}
    \caption{Comparison of the standard landmark annotation (\emph{red} dots) with the curve annotation (\emph{blue} lines) on hands, ears and faces.}
    \label{fig:intro}
\end{figure}

Building and fitting Statistical Deformable Models (SDMs) of various object classes is a well-studied and popular area in the intersection of computer vision and machine learning \cite{Cootes1995, Cootes2001, Matthews2004, Saragih2011, Belhumeur2011, Zhu2012, Xiong2013}. Recently, we have witnessed tremendous developments in building and fitting SDMs of human faces and bodies using images that are captured under unconstrained conditions, usually referred to as ``in-the-wild" \cite{Belhumeur2011, Cao2012, Zhu2012, Xiong2013, Asthana2013, Tzimiropoulos2014, Asthana2014}. This is attributed to:
\begin{itemize}
\item The abundance of complex visual data, spread mostly through the internet via web services such as Youtube, Flickr and Google Images. The latter has led to the development of large ``in-the-wild" databases of human faces and bodies \cite{Belhumeur2011, Le2012, Zhu2012, Burgos2013}.

\item The development of powerful visual features, able to describe objects and its parts in a robust manner (e.g., Scale Invariant Feature Transforms (SIFT) \cite{lowe1999object}, Histogram of Oriented Gradients (HoGs) \cite{Dalal2005} and recently Deep Convolutional Neural Networks \cite{sermanet2013overfeat}), as well as generative and discriminative methodologies for learning deformable models.

\item Equally important is the task of manual annotation of images with regards to face and body parts and that has been undertaken by several research teams \cite{sagonas_iccv_300w_2013,charles2013domain,dantone2014body,andriluka14cvpr}.
\end{itemize}

There are two drawbacks on building deformable models that directly rely on landmarks:
\begin{itemize}

\item Annotating with regards to consistent landmarks is an extremely time consuming, tedious and labour intensive work \cite{sagonas_iccv_300w_2013}, which is usually performed by a trained person. Furthermore, for many objects it requires a highly skilled person to identify and annotate a set of images with regards to a consistent set of landmarks. For example human ears have very complicated inner structures (e.g. helix, crus antihelicis, scapha, tragus, lobe etc.) which differ remarkably between ears. Moreover, certain ear parts, such as  fossa triangularies and crus helicis, do not appear in all ears and  their visibility is highly sensitive to pose and illumination variation. Another example is human body which is generally annotated with regards to a number of landmarks that intuitively correspond to a set of body joints. The task of annotation was undertaken by a crowd-sourcing Internet marketplace, so-called Amazon Mechanical Turk (AMT). Unfortunately, these annotations are, in many cases, not consistent\footnote{In the case of faces, the quality of annotations produced from AMT are extremely inaccurate and cannot by any means be compared with the ones provided by the recent competition \cite{sagonas_iccv_300w_2013}} (please see Figure \ref{fig:wrong_anno}). As it was also pointed out recently \cite{tompson2015efficient}, the inconsistencies in body joint annotations may also render the comparison between different human pose estimation methodologies irrelevant.

\item The nature of many deformable objects does not allow them to be annotated with regards to a consistent set of landmarks (e.g.~bottles, fruits etc.). Also the boundary/outline of objects, such as faces, ears, body can be very difficultly annotated consistently. That is why many state-of-the-art methods opt to leave the boundary out from the comparisons \cite{Tzimiropoulos2014, Asthana2014}. The majority of the state-of-the-art methods for model-based landmark localisation \cite{Cao2012, Zhu2012, Xiong2013, Tzimiropoulos2014, Asthana2014} are not applicable for objects with inconsistent sets of landmarks.

@@

To illustrate how time consuming careful annotation of the human ear is we lay down our own experience. A trained annotator needs an average of 4 minutes per image for the manual annotation of 55 landmarks. This means that the annotation of 1000 images requires a total of about 67 hours. Furthermore,  training and fatigue largely influence the annotation accuracy. Hence, a second pass on the annotated data is, in many cases, necessary. Due to the fact that manual annotation is a costly and labour-intensive procedure, unsupervised learning of deformable models for the tasks of object alignment has recently attracted some attention \cite{frey2003learning, baker2004automatic, cootes2004groupwise, jojic2006escaping, Huang2006, kokkinos2007unsupervised, jiang2009learning, liu2009simultaneous, Zhang2012}. However, because the problem of fully unsupervised discovery of the deformations of arbitrary objects is difficult and ill-posed the limited number of methods that have been applied to the task cannot be directly applied to arbitrary collection of images collected ``in-the-wild". On the other hand, the method of \cite{antonakos2014automatic}, which can deal with ``in-the-wild" images, requires a set of consistently annotated sparse shapes to perform deformable face congealing.

@@

In this paper, we provide a solution for annotating an object with regards to its deformations that requires considerable less effort and in the same time can be used to define statistical deformations for objects without consistent set of landmarks. We apply the proposed method to construct deformable models of the outline of  human body parts (e.g., arms and legs). To the best of our knowledge, this is the first SDM of the outline of human body parts. The proposed SDM can be also used to provide accurate and consistent annotations of several of the body joints (such as wrist etc.).

To this end, we argue and empirically demonstrate that it is better to annotate an object with regards to a set of continuous lines that describe the shape of the object. An example is provided in Figure \ref{fig:intro}, where classical landmark annotations are compared with the proposed curve annotations for hands, ears and faces. It is evident that curve annotations avoid the inherent ambiguity of placing sparse landmarks and offer a richer description of the object shape.
%
%where the shape of the human arm is annotated with continuous lines instead of a number of discrete landmarks. Similarly, the shape annotations for an ear and a face are shown in the same figure. Annotating images by following this procedure takes 42, 38 and 43 secs per image for arm, faces, ears correspondingly.
Furthermore, these curves can be automatically produced by recent methods that perform discriminative segmentation of objects.

In addition, we capitalise on recent advances on multiframe optical flow estimation \cite{Garg:2013hu} and show that the relevant methodologies have matured enough to densely annotate the proposed shapes using simple, as well as more recent and robust shape representations methods \cite{Nguyen2013}.
In particular, to build dense correspondences between different shapes of the same object type, we estimate the optical flow among all shapes jointly by imposing low-rank constrains, an approach that we call \emph{Shape Flow}. Multiframe optical flow is originally applied on video sequences, relying on the assumptions of color constancy and motion smoothness. However, these assumptions do not hold in our case, where we have a collection of shapes, therefore we introduce appropriate modifications of the multiframe optical flow formulation, based on consistency of the image-based shape representation and low rank priors.
%
%On the other hand,  while shapes having no motion smoothness completely as all training data are independent. Low rank constrain plays important role to simulate motion smoothness assumption.

After establishing dense shape correspondences, we describe how dense Active Appearance Models (AAMs) can be built on top of that. We show that the proposed methodology can be applied for landmark localisation, even though not tailored for that task, achieving particularly good performance. We demonstrate the performance of the proposed methodology for describing objects that do not have consistent landmarks. Furthermore, We show that by sampling the shape of the dense AAM, a part-based AAM can be trained with its parts being in correspondence. We show that such a part-based AAM can used to produce state-of-the-art results in localisation of human body parts in challenging databases, outperforming state-of-the-art methods \cite{pfister2015flowing}. Finally, we show that the proposed part-based AAM can be used to provide consistent annotations for different body parts.

>>>>

\begin{figure}[t!]
    \centering
    \begin{subfigure}[b]{0.09125\textwidth}
            \includegraphics[height=3.25cm]{resources/MotivativeAnnotation/MPII}
            \caption{MPII}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.1125\textwidth}
            \includegraphics[height=3.25cm]{resources/MotivativeAnnotation/FashionPose}
            \caption{Fashion}
    \end{subfigure}
  	\hfill
    \begin{subfigure}[b]{0.125\textwidth}
            \includegraphics[height=3.25cm]{resources/MotivativeAnnotation/FLIC}
            \caption{FLIC}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.125\textwidth}
            \includegraphics[height=3.25cm]{resources/MotivativeAnnotation/BBCPose}
            \caption{BBC Pose}
    \end{subfigure}
    \caption{Examples of inconsistent and inaccurate annotations of human pose among different datasets. \emph{Blue} dots denote the original annotations. The arrows and \emph{green} dots show the correct location at which the points should be.
    %Original annotations are shown in blue dots while better annotations shown in green dots.
    }
    \label{fig:wrong_anno}
\end{figure}
\begin{figure}[t!]
    \centering
    \begin{subfigure}[b]{0.065\textwidth}
            \includegraphics[height=2.55cm]{resources/Fig_Intro/intro_2_0}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.065\textwidth}
            \includegraphics[height=2.55cm]{resources/Fig_Intro/intro_2_1}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.065\textwidth}
            \includegraphics[height=2.55cm]{resources/Fig_Intro/intro_2_2}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.085\textwidth}
            \includegraphics[height=2.55cm]{resources/Fig_Intro/intro_1_0}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.085\textwidth}
            \includegraphics[height=2.55cm]{resources/Fig_Intro/intro_1_1}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.085\textwidth}
            \includegraphics[height=2.55cm]{resources/Fig_Intro/intro_1_2}
    \end{subfigure}
    \\
    \begin{subfigure}[b]{0.156\textwidth}
            \includegraphics[height=2.8cm]{resources/Fig_Intro/intro_0_0}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.156\textwidth}
            \includegraphics[height=2.8cm]{resources/Fig_Intro/intro_0_1}
    \end{subfigure}
  	\hfill
    \begin{subfigure}[b]{0.156\textwidth}
            \includegraphics[height=2.8cm]{resources/Fig_Intro/intro_0_2}
    \end{subfigure}
    \caption{Comparison of the standard landmark annotation (\emph{red} dots) with the curve annotation (\emph{blue} lines) on arms, ears and faces. It is evident that the curve annotations surpass the inevitable inconsistency of sparse annotations.}
    \label{fig:intro}
\end{figure}

Statistical Deformable Models (SDMs) of various object classes is a well-studied and popular area in the intersection of computer vision and machine learning \cite{Cootes1995, Cootes2001, Matthews2004, Saragih2011, Belhumeur2011, Zhu2012, Xiong2013}. Recently, we have witnessed tremendous developments on SDMs of human faces and bodies trained with images that are captured under unconstrained conditions, usually referred to as ``in-the-wild" \cite{Belhumeur2011, Cao2012, Zhu2012, Xiong2013, Asthana2013, Tzimiropoulos2014, Asthana2014}. This is attributed to:
\begin{itemize}
\item The abundance of complex visual data, spread through web services (e.g. Youtube, Flickr, Google Images), which has led to the development of ``in-the-wild" databases of human faces and bodies \cite{Belhumeur2011, Le2012, Zhu2012, Burgos2013}.

\item The manual annotation of such databases that has been undertaken by several research teams \cite{sagonas_iccv_300w_2013,charles2013domain,dantone2014body,andriluka14cvpr}.

\item The development of powerful visual features that are able to describe objects and their parts in a robust manner (e.g., SIFT \cite{lowe1999object}, HoGs \cite{Dalal2005}
%and Deep Convolutional Neural Networks \cite{sermanet2013overfeat}
), as well as generative and discriminative methodologies for learning SDMs.
\end{itemize}

However, there are two main drawbacks when building SDMs directly on manually annotated landmarks:
\begin{itemize}

\item Annotating with regards to consistent landmarks is an extremely time consuming, tedious and labour intensive work \cite{sagonas_iccv_300w_2013}, which is usually performed by a trained person. Furthermore, for various object classes, it requires a highly skilled person in order to identify and annotate landmarks in a consistent manner. For example, the human ear has very complicated inner structures (helix, crus antihelicis, scapha, tragus, lobe etc.) which remarkably differ between different ears. Moreover, certain ear parts, such as fossa triangularies and crus helicis, do not appear in all ears and their visibility is highly sensitive to the head pose and illumination variation. Another such example is the human body, which is generally annotated with regards to a number of landmarks that intuitively correspond to a set of body joints. For most body pose databases, annotation task was undertaken by a crowd-sourcing Internet marketplace, so-called Amazon Mechanical Turk (AMT). Unfortunately, this resulted in acquiring inconsistent and inaccurate annotations, in many cases\footnote{In the case of faces, the quality of annotations produced from AMT are extremely inaccurate and cannot, by any means, be compared with the ones provided by the recent 300W competition \cite{sagonas_iccv_300w_2013}.} (please see Figure \ref{fig:wrong_anno}). As it was also recently pointed out \cite{tompson2015efficient}, the inconsistencies in body joint annotations may also render the comparison between different human pose estimation methodologies irrelevant.

\item The nature of many deformable objects does not allow them to be annotated with regards to a consistent set of landmarks (e.g., bottles, fruits etc.). Additionally, it is very difficult to consistently annotate the boundary/outline of certain objects such as faces, ears, body, since these landmarks do not have any semantic meaning. That is why many state-of-the-art methods opt to leave the boundary out when reporting results \cite{Tzimiropoulos2014, Asthana2014}. The majority of the state-of-the-art methods for model-based landmark localisation \cite{Cao2012, Zhu2012, Xiong2013, Tzimiropoulos2014, Asthana2014} are not applicable to objects with inconsistent sets of landmarks.

@@

To illustrate how time consuming careful annotation of a complex deformable object is, we lay down our own experience based on the human ear. A trained annotator needs an average of 4 minutes per image for the manual annotation of 55 landmarks. This means that the annotation of 1000 images requires a total of about 67 hours. Furthermore, the quality of training as well as fatigue largely influence the annotation accuracy. Hence, a second pass on the annotated data is, in many cases, necessary. Due to the fact that manual annotation is a costly and labour-intensive procedure, unsupervised learning of deformable models for the task of object alignment has recently attracted some attention \cite{frey2003learning, baker2004automatic, cootes2004groupwise, jojic2006escaping, Huang2006, kokkinos2007unsupervised, jiang2009learning, liu2009simultaneous, Zhang2012}. However, because the problem of fully unsupervised discovery of the deformations of arbitrary objects is difficult and ill-posed, the limited number of methods that have been proposed for the task cannot be directly applied to arbitrary collection of ``in-the-wild" images. On the other hand, the method of \cite{antonakos2014automatic}, which can deal with ``in-the-wild" images, requires a set of consistently annotated sparse shapes to perform deformable face congealing.

@@

In this paper, we propose a solution for annotating an object with regards to its deformations that requires considerably less effort compared to manual annotation and, at the same time, can be used to define statistical deformations for objects without a consistent set of landmarks. We employ the proposed method in order to construct SDMs based on the outline of human body parts (i.e., arms and legs). To the best of our knowledge, this is the first time that an SDM is proposed to be trained on the outline of human body parts and not on their skeleton joints. The proposed SDM can also be used to provide accurate and consistent annotations for several of the body joints (such as wrist, elbow etc.).

To this end, we argue and empirically demonstrate that it is better to annotate an object with regards to a set of continuous lines that describe its shape. An example is provided in Figure \ref{fig:intro}, which compares the standard landmark annotations that are employed in the current literature with the proposed curve annotations for arms, ears and faces. It becomes evident that the curve annotations avoid the inherent ambiguity of placing sparse landmarks and offer a richer description of the object's shape.
%
%where the shape of the human arm is annotated with continuous lines instead of a number of discrete landmarks. Similarly, the shape annotations for an ear and a face are shown in the same figure. Annotating images by following this procedure takes 42, 38 and 43 secs per image for arm, faces, ears correspondingly.
Furthermore, these curves can be automatically generated by recently proposed methods that perform discriminative segmentation of objects \cite{luo2013pedestrian,liu2015matching}.

Furthermore, we capitalise on recent advances on multiframe optical flow estimation \cite{Garg:2013hu} and show that the relevant methodologies have matured enough to densely annotate the proposed shapes using either simplistic or more even sophisticated and robust shape representation methods \cite{Nguyen2013}.
In particular, to build dense correspondences between different shape instances of the same object class, we jointly estimate the optical flow among all the instances by imposing low-rank constrains, an approach that we call \emph{Shape Flow}. Multiframe optical flow has originally been applied on video sequences, relying on the assumptions of color consistency and motion smoothness \cite{Garg:2013hu}. However, these assumptions do not hold in our case, where we have a collection of shapes. Therefore, we introduce appropriate modifications based on the consistency of image-based shape representation, as well as low rank priors.
%
%On the other hand,  while shapes having no motion smoothness completely as all training data are independent. Low rank constrain plays important role to simulate motion smoothness assumption.

In addition, we demonstrate how dense Active Appearance Models (AAMs) can be built after establishing dense shape correspondences. We show that the proposed methodology can be applied for landmark localisation, even though it is not tailored for that task, achieving particularly good performance. We also illustrate the performance of the proposed methodology for describing objects that do not have consistent landmarks. We report state-of-the-art

Furthermore, we show that by sampling the shape of the dense AAM, a part-based AAM can be trained with its parts being in correspondence. We show that such a part-based AAM can used to produce state-of-the-art results in localisation of human body parts in challenging databases, outperforming state-of-the-art methods \cite{pfister2015flowing}. Finally, we show that the proposed part-based AAM can be used to provide consistent annotations for different body parts.
